$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-06-22 paper/paper.tex
Add work by Ciccone and Padovani

Allais \cite{Allais2018a} uses leftover typing to mechanise in Agda a bidirectional type system for the linear \lambdacalc{}.
He proves type preservation and provides a decision procedure for type checking and type inference.
In our paper, we apply leftover typing to the \picalc{} following Allais \cite{Allais2018a}, and focus on type checking. We give a more general usage algebra, leading to linear, graded and shared type systems.
Drawing from quantitative type theory (by McBride and Atkey \cite{McBride2016, Atkey2018}), in our work, we too are able to talk about fully consumed resources.
As such, we can transmit over a channel the $\lz$ multiplicities of a fully exhausted channel.
%
Orchard et al. introduce Granule \cite{Orchard}, a fully-fledged functional language with graded modal types, linear types, indexed types and polymorphism.
Modalities include exact usages, security levels and intervals; resource algebras are pre-ordered semirings with partial addition.
They provide bidirectional typing rules, and show the type safety of their semantics.

Recent years have seen an increase in the efforts to mechanise resource-aware process algebras.
One of the earliest works is the mechanisation of the linear \picalc{} in Isabelle/HOL by Gay \cite{Gay2001}.
Gay encodes the \picalc{} with linear and shared types using de Bruijn indices, a reduction relation and a type system posterior to the syntax.
However, in his work context splits are extrinsic, and the types of used multiplicities erased.
We make context splits intrinsic, preserve the ability to talk about used resources, and adopt a more general usage algebra: our framework integrates three different type systems into one.
%
The work by Goto et al. \cite{Goto2016a} is, to the best of our knowledge, the first formalisation of session types that includes a type safety proof (in Coq). 
The authors extend session types with polymorphism and pattern matching.
They use a locally-nameless encoding for variable references, a syntax prior to types, and an LTS semantics that encodes session-typed processes into the \picalc{}.
Their type system uses reordering of contexts and extrinsic context splits. 
%
Castro et al. \cite{Castro2020} provide tooling for locally-nameless representations of process calculi in Coq  --- in Coq de Bruijn indices are not as popular as in Agda or Idris because dependent pattern matching is far from straightforward.
As a use-case, they use their tool to help automate proofs of subject reduction for a type system with session types.
%
Taking an intrinsic approach to typing, Thiemann formalises in Agda the MicroSession (minimal GV \cite{Gay2010}) calculus with support for recursion and subtyping \cite{Thiemann2019}.
Context splits are still given extrinsically, and exhausted resources are removed from typing contexts altogether.
The semantics are given as an intrinsically typed CEK machine with a global context of session-typed channels.
%
Most recently, Rouvoet et al. provide an intrinsic type system for a \lambdacalc{} with session types \cite{Rouvoet2020}.
They use proof relevant separation logic and a notion of a supply and demand \emph{market} to make context splits transparent to the user.
Their separation logic is based on a partial commutative monoid that need not be deterministic nor cancellative.
Their typing rules preserve the balance between supply and demand, and are extremely elegant.
They distill their typing rules even further by modelling the supply and demand market as a state monad.
%
Based on contextual type theory \cite{Pientkaa, Pientka}, LINCX \cite{Georges2017} extends the linear logical framework LLF \cite{Cervesato1996} by internalising the notion of bindings and contexts.
The result is a meta-theory in which HOAS encodings with both linear and dependent types can be described.
The developer obtains for free an equational theory of substitution and decidable typechecking without having to encode context splits within the object language.

>>>>

Allais \cite{Allais2018a} uses leftover typing to mechanise in Agda a bidirectional type system for the linear \lambdacalc{}.
He proves type preservation and provides a decision procedure for type checking and type inference.
In our paper, we apply leftover typing to the \picalc{} following Allais \cite{Allais2018a}, and focus on type checking. We give a more general usage algebra, leading to linear, graded and shared type systems.
Drawing from quantitative type theory (by McBride and Atkey \cite{McBride2016, Atkey2018}), in our work, we too are able to talk about fully consumed resources.
As such, we can transmit over a channel the $\lz$ multiplicities of a fully exhausted channel.
%
In their recent paper (unpublished at the time of writing), Ciccone and Padovani mechanise a dependently-typed linear \picalc{} in Agda \cite{Ciccone}.
They do so by defining an intrinsicly-typed calculus, which allows them to leverage Agda to provide a dependently-typed interpretation of messages.
Contexts contain not only types, but also the messages that inhabit their corresponding interpretation in Agda.
To avoid linearity violations, channel types as interpreted as unit types --- and therefore only present in the context in the object language.
This approach allows them to encode message predicates, branching, and variable-length conversations, as message input is modelled as dependent function abstraction in Agda.
They use the concrete set of multiplicities $0$, $1$, $\omega$, define a partial algebra on contexts and require extrinsic proofs of context split.
%
Orchard et al. introduce Granule \cite{Orchard}, a fully-fledged functional language with graded modal types, linear types, indexed types and polymorphism.
Modalities include exact usages, security levels and intervals; resource algebras are pre-ordered semirings with partial addition.
They provide bidirectional typing rules, and show the type safety of their semantics.

Recent years have seen an increase in the efforts to mechanise resource-aware process algebras.
One of the earliest works is the mechanisation of the linear \picalc{} in Isabelle/HOL by Gay \cite{Gay2001}.
Gay encodes the \picalc{} with linear and shared types using de Bruijn indices, a reduction relation and a type system posterior to the syntax.
However, in his work context splits are extrinsic, and used multiplicities erased from context.
We make context splits intrinsic, preserve the ability to talk about used resources, and adopt a more general usage algebra: our framework integrates three different type systems into one.
%
The work by Goto et al. \cite{Goto2016a} is, to the best of our knowledge, the first formalisation of session types that includes a type safety proof (in Coq). 
The authors extend session types with polymorphism and pattern matching.
They use a locally-nameless encoding for variable references, a syntax prior to types, and an LTS semantics that encodes session-typed processes into the \picalc{}.
Their type system uses reordering of contexts and extrinsic context splits. 
%
Castro et al. \cite{Castro2020} provide tooling for locally-nameless representations of process calculi in Coq  --- in Coq de Bruijn indices are not as popular as in Agda or Idris because dependent pattern matching is far from straightforward.
As a use-case, they use their tool to help automate proofs of subject reduction for a type system with session types.
%
Taking an intrinsic approach to typing, Thiemann formalises in Agda the MicroSession (minimal GV \cite{Gay2010}) calculus with support for recursion and subtyping \cite{Thiemann2019}.
Context splits are still given extrinsically, and exhausted resources are removed from typing contexts altogether.
The semantics are given as an intrinsically typed CEK machine with a global context of session-typed channels.
%
Most recently, Rouvoet et al. provide an intrinsic type system for a \lambdacalc{} with session types \cite{Rouvoet2020}.
They use proof relevant separation logic and a notion of a supply and demand \emph{market} to make context splits transparent to the user.
Their separation logic is based on a partial commutative monoid that need not be deterministic nor cancellative.
Their typing rules preserve the balance between supply and demand, and are extremely elegant.
They distill their typing rules even further by modelling the supply and demand market as a state monad.
%
Based on contextual type theory \cite{Pientkaa, Pientka}, LINCX \cite{Georges2017} extends the linear logical framework LLF \cite{Cervesato1996} by internalising the notion of bindings and contexts.
The result is a meta-theory in which HOAS encodings with both linear and dependent types can be described.
The developer obtains for free an equational theory of substitution and decidable typechecking without having to encode context splits within the object language.
